---
title: "Healthcare_Expenditure_in_aging_era"
author: "Wanting Echo Zhao"
date: "2025-12-27"
output:
  pdf_document: default
  html_document: default
---

## Import libary

```{r echo=TRUE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(plm)
library(ggplot2)
```

## Read Data

```{r echo=TRUE}
ghed_raw <- read_excel("/Users/ez_us/Documents/Aging_HCE/datasets/GHED_data.XLSX", sheet= "Data")
```

### Check the variable names

```{r}
str(ghed_raw)
```

## Select relevant GHED variables

```{r}
ghed_clean <- ghed_raw %>%
  select(
    location,
    code,
    year,
    gghed_che,
    shi_che,
    oops_che,
    vhi_che
  ) %>%
  filter(year >= 2000 & year <= 2023) %>%
  mutate(
    year = as.integer(year)
  )
```

### Sanity check

```{r}
summary(ghed_clean)
colSums(is.na(ghed_clean))
```

## Read World Bank aging data

```{r}
aging_raw <- read_excel(
  "/Users/ez_us/Documents/Aging_HCE/datasets/API_SP.POP.65UP.TO.ZS_DS2_en_excel_v2_1001.xls",
  sheet = "Data",
  skip = 3
)
```

## Reshape Wolrd Bank data (wide to long)

```{r}
aging_long <- aging_raw %>%
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "aging_65"
  ) %>%
  rename(
    location = `Country Name`,
    code = `Country Code`
  ) %>%
  mutate(
    year = as.integer(year)
  ) %>%
  filter(year >= 2000 & year <= 2023)
```

### Check

```{r}
str(aging_long)
summary(aging_long$aging_65)
```

### Check for coverage gaps,which helps later when deciding which countries to keep.

```{r}
aging_long %>%
  group_by(location) %>%
  summarise(n_years = n()) %>%
  arrange(n_years)
```

## Merge GHED + World Bank

```{r}
panel_data <- ghed_clean %>%
  left_join(aging_long, by = c("code", "year"))
```

### Check result

```{r}
dim(panel_data)
summary(panel_data)
```

## Convert to panel-data format (for models)

```{r}
panel_data <- pdata.frame(
  panel_data,
  index = c("code", "year")
)
```

## First baseline models (run separately)

### Example 1: Out-of-pocket share

```{r}
model_oops <- plm(
  oops_che ~ aging_65,
  data = panel_data,
  model = "within"
)

summary(model_oops)
```

### Goverment financing share

```{r}
model_gov <- plm(
  gghed_che ~ aging_65,
  data = panel_data,
  model = "within"
)

summary(model_gov)
```

### Social health insuramce

```{r}
model_shi <- plm(
  shi_che ~ aging_65,
  data = panel_data,
  model = "within"
)

summary(model_shi)
```

### Voluntary health insurance

```{r}
model_vhi <- plm(
  vhi_che ~ aging_65,
  data = panel_data,
  model = "within"
)

summary(model_vhi)
```

## Asia-focused descriptive check(bonus)

```{r}
asia_codes <- c("CHN", "JPN", "KOR", "THA")

asia_data <- panel_data %>%
  filter(code %in% asia_codes) %>%
  as.data.frame()

asia_data <- pdata.frame(
  asia_data,
  index = c("code", "year")
)
```

## test - generate oop plot

```{r}
ggplot(asia_data, aes(x = year, y = oops_che, color = code, group = code)) + 
  geom_line() + 
  labs(
  title = "Out-of-pocket Health Expenditure in Aging Asia Economies",
  x = "Year",
  y = "% of Current Health Expenditure"
)
```

## Run financing models again

### Example 1: Out-of-pocket share

```{r}
model_asia_oops <- plm(
  oops_che ~ aging_65,
  data = asia_data,
  model = "within"
)

summary(model_asia_oops)
```

#### Interpretation

> Holding country-specific characteristics constant, a one-percentage-point increase in the population aged 65+ is associated with a 1.5 percentage-point decline in the out-of-pocket share of health expenditure in Asian economies.

### Example 2: Government financing share

```{r}
model_asia_gov <- plm(
  gghed_che ~ aging_65,
  data = asia_data,
  model = "within"
)

summary(model_asia_gov)
```

### Interpretation

> Aging is associated with a statistically significant increase in the government share of health expenditure.

> This indicates: - Fiscal absorption of aging-related costs - Active public policy response - Alignment with UHC expansion trajectories

This result explains where the OOP reduction goes. \### Example 3: Social health insurance

```{r}
model_asia_shi <- plm(
  shi_che ~ aging_65,
  data = asia_data,
  model = "within"
)

summary(model_asia_shi)
```

### Example 4: Voluntary health insurance

```{r}
model_asia_vhi <- plm(
  vhi_che ~ aging_65,
  data = asia_data,
  model = "within"
)

summary(model_asia_vhi)
```

```{r}
ggplot(asia_data, aes(x = year, y = gghed_che, color = code, group = code)) + 
  geom_line() + 
  labs(
  title = "Domestic General Government Health Expenditure in Aging Asia Economies",
  x = "Year",
  y = "% of Current Health Expenditure"
)

ggplot(asia_data, aes(x = year, y = shi_che, color = code, group = code)) + 
  geom_line() + 
  labs(
  title = "Social Health Insurance in Aging Asia Economies",
  x = "Year",
  y = "% of Current Health Expenditure"
)
```

# Core question: “Is the effect stronger in countries that are actually aging rapidly?”

### This tests whether your results are mechanically driven by aging, not by global trends.

## Calculate country-level aging change

```{r}
aging_intensity <- panel_data %>%
  filter(year %in% c(2000, 2023)) %>%
  select(code, year, aging_65) %>%
  group_by(code) %>%
  filter(n() == 2) %>%              # KEEP only countries with both years
  ungroup() %>%
  pivot_wider(
    names_from = year,
    values_from = aging_65
  ) %>%
  mutate(
    aging_change = `2023` - `2000`
  )
```

## High-aging vs low-aging (median split)

```{r}
median_change <- median(aging_intensity$aging_change, na.rm = TRUE)

aging_intensity <- aging_intensity %>%
  mutate(
    high_aging = ifelse(aging_change >= median_change, 1, 0)
  )

panel_data <- as.data.frame(panel_data)

panel_data <- panel_data %>%
  mutate(code = as.character(code)) %>%   # ensure same type
  left_join(
    aging_intensity %>%
      mutate(code = as.character(code)) %>%
      select(code, aging_change, high_aging),
    by = "code"
  )
  
panel_data <- pdata.frame(
  panel_data,
  index = c("code", "year")
)
```

## Robustness model with interaction (core result)

```{r}
model_oops_robust <- plm(
  oops_che ~ aging_65 * high_aging,
  data = panel_data,
  model = "within",
  index = c("code", "year")
)

summary(model_oops_robust)
```

```{r}
model_gghed_robust <- plm(
  gghed_che ~ aging_65 * high_aging,
  data = panel_data,
  model = "within",
  index = c("code", "year")
)

summary(model_gghed_robust)
```

```{r}
model_shi_robust <- plm(
  shi_che ~ aging_65 * high_aging,
  data = panel_data,
  model = "within",
  index = c("code", "year")
)

summary(model_shi_robust)
```

```{r}
library(stargazer)

stargazer(
  model_oops_robust,
  model_gghed_robust,
  model_shi_robust,
  type = "text",
  title = "Interaction Effects of Population Aging on Health Financing Structure",
  dep.var.labels = c("Out-of-pocket (%)", "Government (%)", "Social Health Insurance (%)"),
  covariate.labels = c("Aging 65+", "Aging 65+ × High-aging"),
  omit.stat = c("rsq", "adj.rsq", "f")
)
```

Across countries, increases in the elderly population are associated with a consistent shift away from out-of-pocket healthcare spending toward government and insurance-based financing. Importantly, this shift is significantly stronger in countries experiencing faster demographic aging, suggesting that health systems actively redistribute financial responsibility as aging pressure intensifies.

## Create financing components(group oop, gghed, shi, and vhi)

```{r}
hero_data <- panel_data %>%
  as.data.frame() %>%
  filter(!is.na(high_aging)) %>%
  select(
    year,
    high_aging,
    oops_che,
    gghed_che,
    shi_che,
    vhi_che
  )
```

## Aggregate by aging intensity and year

```{r}
hero_agg <- hero_data %>%
  group_by(high_aging, year) %>%
  summarise(
    Out_of_Pocket = mean(oops_che, na.rm = TRUE),
    Government = mean(gghed_che, na.rm = TRUE),
    Social_Insurance = mean(shi_che, na.rm = TRUE),
    Voluntary_Insurance = mean(vhi_che, na.rm = TRUE),
    .groups = "drop"
  )
```

## Reshape to long format for ggplot

```{r}
hero_long <- hero_agg %>%
  as.data.frame() %>%   # DROP pdata.frame attributes
  pivot_longer(
    cols = c(
      Out_of_Pocket,
      Government,
      Social_Insurance,
      Voluntary_Insurance
    ),
    names_to = "Financing_Type",
    values_to = "Share"
  ) %>%
  mutate(
    Aging_Group = ifelse(
      high_aging == 1,
      "High-aging countries",
      "Low-aging countries"
    ),
    Financing_Type = recode(
      Financing_Type,
      "Out_of_Pocket" = "Out-of-pocket payments",
      "Government" = "Government financing",
      "Social_Insurance" = "Social health insurance",
      "Voluntary_Insurance" = "Voluntary health insurance"
    )
  )
```

## 

```{r}
unique(hero_long$Financing_Type)
levels(hero_long$Financing_Type)
str(hero_long$Financing_Type)

ggplot(hero_long, aes(x = year, y = Share, fill = Financing_Type)) +
  geom_area(alpha = 0.9, color = "white", linewidth = 0.2) +
  facet_wrap(~ Aging_Group) +
  scale_fill_manual(
    values = c(
      "Out-of-pocket payments" = "#D55E00",
      "Government financing" = "#0072B2",
      "Social health insurance" = "#009E73",
      "Voluntary health insurance" = "#CC79A7"
    )
  ) +
  labs(
    title = "How Population Aging Reshapes Healthcare Financing",
    subtitle = "Average composition of health expenditure in high- and low-aging countries (2000–2023)",
    x = "Year",
    y = "Share of total health expenditure (%)",
    fill = "Financing source",
    caption = "Source: WHO GHED; World Bank. Countries grouped by aging intensity (2000–2023)."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
